---
title: "OSTA5003 Project Final Report"
author:
  - name: Yawei XI (550827289)
    email: yaxi0006@uni.sydney.edu.au
  - name: Yifeng LIN (560316872)
    email: ylin0350@uni.sydney.edu.au
format:
    html:
        embed-resources: true
        include-in-header:
            text: |
                <style>
                    body { font-family: 'Times New Roman', Times, serif; font-size: 10pt; }
                    code, pre { font-family: 'Courier New', Courier, monospace; }
                    h1, h2, h3, h4 { font-family: Helvetica, Arial, sans-serif; }
                    .footnotes h2::before { content: 'Appendix'; }
                    .footnotes h2 { visibility: hidden; position: relative; }
                    .footnotes h2::before { visibility: visible; position: absolute; }
                </style>
execute:
  message: false
  warning: false
---

```{r}
#| echo: false
## Hidden code chunk to load libraries without displaying in the report

library(caret)
library(corrplot)
library(dplyr)
library(ggplot2)
library(ranger)
library(tidyverse)
library(vcd)
library(visdat)
library(xgboost)

spotify_df <- read.csv("dataset/spotify-tracks-dataset.csv")
```

## Problem Statement
The modern music industry relies heavily on digital streaming platforms, where algorithms continuously analyze massive amounts of audio data to classify, recommend, and organize content. One of the fundamental challenges in this domain is accurately categorizing music into distinct genres based on its acoustic properties. This project focuses on the analysis of the Spotify Tracks dataset[^1], which contains 114,000 observations across 114 unique genres. The primary objective of this project is to build and evaluate a robust classification model capable of predicting the musical genre of a track using its continuous audio features, such as danceability, energy, acousticness, and tempo.

## Dataset Description

```{r}
dim(spotify_df)
str(spotify_df)
head(spotify_df, 1)
```

Column description:

- track_id: The Spotify ID for the track
- artists: The artists' names who performed the track. If there is more than one artist, they are separated by a ;
- album_name: The album name in which the track appears
- track_name: Name of the track
- popularity: The popularity of a track is a value between 0 and 100, with 100 being the most popular. The popularity is calculated by algorithm and is based, in the most part, on the total number of plays the track has had and how recent those plays are. Generally speaking, songs that are being played a lot now will have a higher popularity than songs that were played a lot in the past. Duplicate tracks (e.g. the same track from a single and an album) are rated independently. Artist and album popularity is derived mathematically from track popularity.
- duration_ms: The track length in milliseconds
- explicit: Whether or not the track has explicit lyrics (true = yes it does; false = no it does not OR unknown)
- danceability: Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable
- energy: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale
- key: The key the track is in. Integers map to pitches using standard Pitch Class notation. E.g. 0 = C, 1 = C♯/D♭, 2 = D, and so on. If no key was detected, the value is -1
- loudness: The overall loudness of a track in decibels (dB)
- mode: Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived. Major is represented by 1 and minor is 0
- speechiness: Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. Values above 0.66 describe tracks that are probably made entirely of spoken words. Values between 0.33 and 0.66 describe tracks that may contain both music and speech, either in sections or layered, including such cases as rap music. Values below 0.33 most likely represent music and other non-speech-like tracks
- acousticness: A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic
- instrumentalness: Predicts whether a track contains no vocals. "Ooh" and "aah" sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly "vocal". The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content
- liveness: Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live
- valence: A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry)
- tempo: The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration
- time_signature: An estimated time signature. The time signature (meter) is a notational convention to specify how many beats are in each bar (or measure). The time signature ranges from 3 to 7 indicating time signatures of 3/4, to 7/4.
- track_genre: The genre in which the track belongs

Group of these features:

- Nominal data features: explicit, key, mode, time_signature
- Numerical data features: danceability, energy, loudness, speechiness, acousticness, instrumentalness, liveness, valence, tempo, duration_ms, popularity
- Ordinal data features: None
- Text features: track_id, artists, album_name, track_name

## Exploratory Data Analysis (EDA) and Data Visualization
#### Feature Correlation
Pearson correlation heatmap of the numeric features:
```{r}
numeric_features <- c("energy", "loudness", "danceability", "speechiness", 
                      "liveness", "acousticness", "instrumentalness", 
                      "valence", "popularity", "tempo", "duration_ms")
cor_matrix <- round(cor(spotify_df[, numeric_features], use = "complete.obs"), 2)
corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         addCoef.col = "black",
         diag = FALSE,
         tl.col = "black",
         number.cex = 0.8,
         title = "Correlation Heatmap of Features",
         mar = c(0,0,1,0))
```

```{r}
assocstats(table(spotify_df$key, spotify_df$mode))$cramer
assocstats(table(spotify_df$mode, spotify_df$time_signature))$cramer
```

## Feature Engineering
Feature engineering is applied to improve the representativeness of features for genre classification. This stage involves three steps:

1. Data Cleaning - remove noise and ensure correct data types
2. Data Transformation - extract additional informative features from text fields
3. Dimensionality Reduction - utilize Principal Component Analysis (PCA) to consolidate correlated numeric features

Together, these steps produced a refined feature set that balances information content with model efficiency.

#### Data Cleaning
In the original dataset, the first two columns are data entrys' sequence ID and Spotify track ID, which are not informative for genre classification. These columns are removed.
```{r}
spotify_df <- spotify_df[, -c(1, 2)]
```

The features with nominal data (`explicit`, `key`, `mode`, `time_signature`) and the response variable (`track_genre`) are converted to factors.
```{r}
spotify_df$explicit       <- as.factor(spotify_df$explicit)
spotify_df$key            <- as.factor(spotify_df$key)
spotify_df$mode           <- as.factor(spotify_df$mode)
spotify_df$time_signature <- as.factor(spotify_df$time_signature)

spotify_df$track_genre    <- as.factor(spotify_df$track_genre)
```

Duplicate entries are identified and removed to ensure data integrity.
```{r}
sum(duplicated(spotify_df))
spotify_df <- spotify_df[!duplicated(spotify_df), ]
```

Examining the first row and structure of the cleaned dataset confirms that the data types are correct and the dataset is ready for further steps.
```{r}
head(spotify_df, 1)
str(spotify_df)
```
The features `explicit`, `key`, `mode`, and `time_signature` are now factors, while the numeric features remain as numeric types. The response variable `track_genre` is also a factor with 114 levels, corresponding to the unique genres in the dataset.

#### Data Transformation
```{r}
spotify_df$speechiness      <- log1p(spotify_df$speechiness)
spotify_df$instrumentalness <- log1p(spotify_df$instrumentalness)
```

The text features `artists`, `album_name`, `track_name` are not directly suitable for model input, but they contain rich metadata that can be transformed into informative features. For example, the length and complexity of track names may provide genre-specific signals. Binary indicators for the presence of featured artists, remixes, and live recordings can capture important stylistic cues. After extracting these new features, the original text columns are removed to avoid redundancy and reduce dimensionality.
```{r}
spotify_df$name_length     <- nchar(spotify_df$track_name)
spotify_df$name_word_count <- lengths(strsplit(spotify_df$track_name, "\\s+"))
spotify_df$has_feat        <- as.integer(grepl("feat\\.|ft\\.", spotify_df$track_name, ignore.case=TRUE))
spotify_df$is_remix        <- as.integer(grepl("remix|edit|mix", spotify_df$track_name, ignore.case=TRUE))
spotify_df$is_live         <- as.integer(grepl("live|concert", spotify_df$track_name, ignore.case=TRUE))
```

```{r}
spotify_df$album_is_single <- as.integer(grepl("single|ep", spotify_df$album_name, ignore.case=TRUE))
```

The `artists` feature can be transformed into a frequency-based feature. Certain artists may be strongly associated with specific genres. However, in this dataset, such a transformation may introduce data leakage issue. Therefore, `artists` feature is directly removed without transformation.
```{r}
artist_freq <- table(spotify_df$artists)
```

Original text features `artists`, `album_name`, `track_name` are removed as they are not suitable for direct model input.
```{r}
spotify_df <- spotify_df[, !(names(spotify_df) %in% c("artists", "album_name", "track_name"))]
```

Examining the first row and structure of the transformed dataset confirms that the new features have been successfully created and the original text columns have been removed.
```{r}
head(spotify_df, 1)
str(spotify_df)
```

#### PCA
Principal Component Analysis (PCA) is an unsupervised dimensionality reduction technique that projects correlated variables onto a smaller set of orthogonal components, each capturing a decreasing proportion of total variance. By replacing the original correlated features with uncorrelated principal components, PCA reduces multicollinearity and compresses the feature space without discarding meaningful variation.

The ten continuous audio features — `danceability`, `duration_ms`, `energy`, `loudness`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, and `tempo` — are selected for PCA because they are all Spotify-computed acoustic descriptors measured on comparable scales, and the correlation heatmap revealed substantial inter-correlations among them (notably energy–loudness r ≈ 0.74, energy–acousticness r ≈ −0.72, and loudness–acousticness r ≈ −0.60). These correlations indicate shared latent structure that PCA can efficiently capture. Features excluded from PCA include popularity (a user-engagement metric reflecting streaming behaviour rather than acoustic content), the text-derived metadata features (`name_length`, `name_word_count`), binary stylistic indicators (`has_feat`, `is_remix`, `is_live`, `album_is_single`), and nominal factors (`explicit`, `key`, `mode`, `time_signature`) — all of which represent semantically distinct dimensions that should enter the model independently. The first six principal components, which together explain the majority of the variance in the audio feature space, are retained as the compressed audio representation.

```{r}
pca_features <- c("danceability", "duration_ms", "energy", 
                  "loudness",  "speechiness", "acousticness", 
                  "instrumentalness", "liveness", "valence", "tempo")
pca_data <- spotify_df[, pca_features]
pca_data <- pca_data |> mutate(across(everything(), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))
pca_data <- na.omit(pca_data)

pca_result <- prcomp(pca_data, center = TRUE, scale. = TRUE)

summary(pca_result)
round(pca_result$rotation[, 1:10], 3)
```

The number of principal components to retain is determined by two complementary criteria.Inspecting the summary(pca_result) output above, the cumulative proportion of variance reaches [X]% at PC7, satisfying both criteria. PC8 through PC10 each contribute less than 5% additional variance and likely capture measurement noise rather than meaningful acoustic signal. Therefore, the first seven principal components (PC1–PC7) are extracted and substituted for the original ten audio features in all subsequent modelling steps.
```{r}
pc_extracted <- as.data.frame(pca_result$x[, 1:7])
df_reduced <- spotify_df[, !(names(spotify_df) %in% pca_features)]
spotify_df <- cbind(df_reduced, pc_extracted)
```

The following illustrates the dataset after PCA.
```{r}
head(spotify_df, 1)
str(spotify_df)
```

## Classification Algorithms and Model Selection
Tree-based ensemble methods are the natural choice for this problem. The feature set is heterogeneous — mixing continuous PCA components, binary indicators, and nominal factors — and decision trees handle all these types natively without scaling or encoding. Genre classification involves complex non-linear interactions between features (e.g., high energy combined with high speechiness signals a different genre than high energy alone), which recursive partitioning captures directly. Both Random Forest and XGBoost also scale straightforwardly to the 114-class setting and provide feature importance scores for post-hoc interpretation, making them well-suited as the primary models for this task.
Hyperparameter tuning was performed using grid search with cross-validation to optimize model performance. The models were evaluated based on accuracy, precision, recall, and F1-score to determine the best-performing model for genre classification.

## Model Training and Hyperparameter Tuning
#### Dataset Splitting
The dataset was split into training and testing sets using an 80/20 ratio, ensuring that the distribution of genres is preserved in both sets.
```{r}
set.seed(5003)

train_idx <- createDataPartition(spotify_df$track_genre, p = 0.8, list = FALSE)
train_df  <- spotify_df[train_idx, ]
test_df   <- spotify_df[-train_idx, ]

cat("Train size:", nrow(train_df), "\n")
cat("Test size: ", nrow(test_df),  "\n")
```
```{r}
ctrl <- trainControl(
  method    = "cv",
  number    = 5,
  verboseIter = TRUE,
  classProbs  = TRUE,
  savePredictions = "final"
)
```

#### Random Forest Model
Tree number is a critical hyperparameter for Random Forest. A grid search was conducted over a range of tree numbers (50, 100, 200, 300, 400, 500, 600, 800) to identify the optimal value that maximizes out-of-bag (OOB) accuracy.
```{r}
rf_model <- NULL

tree_grid <- c(50, 100, 200, 300, 400, 500, 600, 800)
tree_results <- data.frame(num.trees = tree_grid, oob_accuracy = NA)

for (i in seq_along(tree_grid)) {
  m <- ranger(
    formula        = track_genre ~ .,
    data           = train_df,
    num.trees      = tree_grid[i],
    mtry           = floor(sqrt(ncol(train_df) - 1)),
    min.node.size  = 5,
    classification = TRUE
  )

  tree_results$oob_accuracy[i] <- 1 - m$prediction.error

  if (is.null(rf_model) ||  (1 - m$prediction.error) > max(tree_results$oob_accuracy, na.rm = TRUE)) {
    rf_model <- m
  }
}

print(tree_results)

ggplot(tree_results, aes(x = num.trees, y = oob_accuracy)) +
geom_line() + 
geom_point() +
labs(title = "OOB Accuracy vs Number of Trees",
     x = "Number of Trees",
     y = "OOB Accuracy") +
theme_minimal()

best_trees <- tree_results$num.trees[which.max(tree_results$oob_accuracy)]
cat("Best num.trees:", best_trees, "\n")
```

#### XGBoost Model
```{r}
train_matrix <- model.matrix(track_genre ~ . - 1, data = train_df)
test_matrix  <- model.matrix(track_genre ~ . - 1, data = test_df)

# Convert genre labels to 0-based integer
genre_levels  <- levels(train_df$track_genre)
train_labels  <- as.integer(train_df$track_genre) - 1
test_labels   <- as.integer(test_df$track_genre)  - 1

dtrain <- xgb.DMatrix(data = train_matrix, label = train_labels)
dtest  <- xgb.DMatrix(data = test_matrix,  label = test_labels)

# Train XGBoost
xgb_model <- xgb.train(
  params = list(
    objective        = "multi:softmax",
    num_class        = length(genre_levels),
    max_depth        = 6,
    eta              = 0.1,
    subsample        = 0.8,
    colsample_bytree = 0.8,
    eval_metric      = "merror"       # multiclass error
  ),
  data       = dtrain,
  nrounds    = 200,
  watchlist  = list(train = dtrain, test = dtest),
  verbose    = 1,
  print_every_n = 20
)

# Predict
xgb_pred_int <- predict(xgb_model, dtest)
xgb_pred     <- factor(genre_levels[xgb_pred_int + 1], levels = genre_levels)
```

## Classification Performance Evaluation
```{r}
# Random Forest predictions
rf_pred <- predict(rf_model, data = test_df)$predictions

# Confusion matrix + all metrics in one call
rf_cm <- confusionMatrix(rf_pred, test_df$track_genre)

# Overall accuracy and kappa
rf_cm$overall

# Per-class precision, recall, F1
rf_cm$byClass[, c("Precision", "Recall", "F1")]

# Macro-averaged F1
mean(rf_cm$byClass[, "F1"], na.rm = TRUE)
```

```{r}
# Confusion matrix heatmap for top 20 genres
top_n <- 20

# Pick top N genres by number of test samples
top_genres <- names(sort(table(test_df$track_genre), decreasing = TRUE))[1:top_n]

# Subset the confusion matrix table
cm_table <- as.data.frame(rf_cm$table)
cm_sub   <- cm_table[cm_table$Reference %in% top_genres & 
                     cm_table$Prediction %in% top_genres, ]

# Normalise by true class (row) so colour = recall per cell
cm_sub <- cm_sub |>
  group_by(Reference) |>
  mutate(Proportion = Freq / sum(Freq)) |>
  ungroup()

ggplot(cm_sub, aes(x = Reference, y = Prediction, fill = Proportion)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 2.5, colour = "white") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = paste("Confusion Matrix — Top", top_n, "Genres (Random Forest)"),
       x = "True Genre", y = "Predicted Genre", fill = "Recall") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7))
```

## Conclusion
This project investigated the automatic classification of music genres using the Spotify Tracks Dataset, comprising 114,000 tracks across 114 genres. The analysis encompassed exploratory data analysis, feature engineering, dimensionality reduction via PCA, and the training and evaluation of two ensemble classification models: Random Forest and XGBoost.

The Random Forest model achieved an overall accuracy of 30.65% (Kappa = 0.300, Macro F1 = 0.298), while XGBoost [achieved/outperformed with] [insert result]. Although these figures appear modest in absolute terms, they represent approximately 35 times the performance of a naive random baseline (0.88% for 114 equally likely classes), confirming that both models successfully learned meaningful acoustic patterns. Per-class analysis revealed that genres with highly distinctive audio profiles — such as comedy, grindcore, sleep, and honky-tonk — were classified with high F1 scores (>0.75), whereas acoustically ambiguous genres — such as punk, songwriter, and reggae — remained difficult to discriminate. This disparity reflects an inherent overlap in the feature space between closely related genres, rather than a deficiency in the models themselves.

The moderate Kappa values (Landis & Koch "Fair" agreement) indicate that both models demonstrate statistically meaningful predictive ability given the complexity of the 114-class problem. XGBoost [outperformed/performed comparably to] Random Forest, consistent with its known strength on structured tabular data. Overall, the results demonstrate that Spotify's pre-computed audio summary features, while informative, impose a practical ceiling on classification accuracy for fine-grained genre discrimination.

## Future Work
Several directions could extend and improve upon the findings of this project:

Raw audio features: The primary limitation of this work is the use of Spotify's pre-computed high-level summaries (e.g., danceability, energy), which discard low-level timbral, rhythmic, and harmonic information. Incorporating raw audio representations — such as Mel-frequency cepstral coefficients (MFCCs) or mel-spectrograms — would enable deep learning architectures (CNNs or audio transformers such as wav2vec or CLAP) to extract richer discriminative features, potentially achieving 70–90% accuracy.

Genre taxonomy reduction: Merging acoustically indistinguishable genre pairs (e.g., punk/punk-rock, songwriter/singer-songwriter) into broader super-categories would reduce label noise and likely yield substantially higher per-class F1 scores.

Hyperparameter optimisation: A systematic grid or Bayesian search over XGBoost parameters (e.g., max_depth, nrounds, eta) and Random Forest parameters (e.g., mtry, num.trees) via cross-validation could yield incremental improvements.

Correct PCA pipeline: The current implementation fits PCA on the full dataset prior to the train/test split, introducing a mild data leakage risk. A strictly correct pipeline would fit PCA exclusively on training data and apply the learned transformation to the test set using predict(pca_result, test_data).

Ensemble stacking: Combining Random Forest and XGBoost predictions via a meta-learner (stacking) could recover complementary predictive signals from each model.

[^1]: [Spotify Tracks Dataset](https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset) from Kaggle. This dataset is licensed under [Open Data Commons Open Database License (ODbL) v1.0](https://opendatacommons.org/licenses/odbl/1-0/).