---
title: "Spotify Tracks Genre Classification Approach"
subtitle: "OSTA5003 Project Presentation"
author:
  - name: Yawei XI (550827289)
    email: yaxi0006@uni.sydney.edu.au
  - name: Yifeng LIN (560316872)
    email: ylin0350@uni.sydney.edu.au
format:
    revealjs:
        auto-slide: 20000
    html:
        embed-resources: true
        include-in-header:
            text: |
                <style>
                    body { font-family: 'Times New Roman', Times, serif; font-size: 10pt; }
                    code, pre { font-family: 'Courier New', Courier, monospace; }
                    h1, h2, h3, h4 { font-family: Helvetica, Arial, sans-serif; }
                    .footnotes h2::before { content: 'Appendix'; }
                    .footnotes h2 { visibility: hidden; position: relative; }
                    .footnotes h2::before { visibility: visible; position: absolute; }
                </style>
execute:
  message: false
  warning: false
---

## 1. Introduction
### 1.1 Background of the Project
Music genre classification is a common machine learning task that aims to predict the category of a song based on its

### 1.2 Phase 2 Objective


### 1.3 Link to Phase 1 Plan

## 2. Data Preparation
### 2.1 Dataset Overview
```{r}
library(tidyverse)
library(caret)
library(ranger)
library(yardstick)
library(ggplot2)

# Load data
spotify_df <- read.csv("dataset/spotify-tracks-dataset.csv")
spotify_df |> dim()
spotify_df |> str()
spotify_df |> head()
```

### 2.2 Feature Selection
```{r}
feature_to_remove <- c("X","track_id", "artists", "album_name", "track_name")
spotify_model <- spotify_df|>
  select(-all_of(feature_to_remove))

spotify_model|>str()
```

### 2.3 Data Cleaning
```{r}
spotify_model <- spotify_model|>
  drop_na()|>
  mutate(
    explicit = as.factor(explicit),
    key = as.factor(key),
    mode = as.factor(mode),
    time_signature = as.factor(time_signature),
    track_genre = as.factor(track_genre)
  )

sum(is.na(spotify_model))
table(spotify_model$track_genre)[1:10]
```


### 2.4 Train-Test Split
```{r}
train_index <- createDataPartition(
  spotify_model$track_genre,
  p = 0.8,
)[[1]]
train_dat <- spotify_model[train_index, ]
test_dat <- spotify_model[-train_index, ]

train_dat|> dim()
test_dat|> dim()
```


## 3. Methodology
### 3.1 Model Choice - Random Forest >>
```{r}
rf_formula <- track_genre ~.
```

### 3.2 Model Training Process >>
```{r}
set.seed(5003)

rf_model<-ranger(
  formula = rf_formula,
  data = train_dat,
  num.trees = 300,
  mtry = floor(sqrt(ncol(train_dat) - 1)),
  min.node.size = 5,
  importance = "impurity",
  classification = TRUE
)

rf_model
```

### 3.3 Hyperparameter Settings >>
```{r}
rf_model$num.trees
rf_model$mtry
rf_model$min.node.size
```


### 3.4 Evaluation Metrics
```{r}
rf_pred <- predict(rf_model, data = test_dat)

test_results <- test_dat|>
  mutate(
    predicted_genre = rf_pred$predictions
  )

head(test_results)
```


## 4. Experimental Results
### 4.1 Overall Performance Metrics >>
```{r}
# Accuracy
accuracy_result <- test_results|>
  accuracy(
    truth = track_genre,
    estimate = predicted_genre
  )

accuracy_result

# F1 score
f1_result <- test_results|>
  f_meas(
    truth = track_genre,
    estimate = predicted_genre,
    estimator = "macro"
  )
f1_result
```

### 4.2 Confusion Matrix >>
```{r}
# Confusion Matrix
cm_result <- test_results|>
  conf_mat(
    truth = track_genre,
    estimate = predicted_genre
  )

cm_result
```

```{r}
# Top 20 genres
top_genres <- test_results|>
  count(track_genre, sort = TRUE)|>
  slice_head(n = 20)|>
  pull(track_genre)

test_top <- test_results|>
  filter(
    track_genre %in% top_genres,
    predicted_genre %in% top_genres
  )

cm_top <- test_top|>
  count(track_genre,predicted_genre)

ggplot(cm_top, aes(x = predicted_genre, y = track_genre, fill = n)) +
  geom_tile() +
  labs(
    title = "Confusion Matrix for Top 20 Genres",
    x = "Predicted Genre",
    y = "Actual Genre",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### 4.3 Class-level Performance
```{r}
```

### 4.4 Feature Importance
```{r}
```

## 5. Discussion
### 5.1 Model Strengthens


### 5.2 Error Analysis from Confusion Matrix >>


### 5.3 Limitaions


### 5.4 Possible Improvements



## 6. Conclusion


## Project Plan
- **Week 1: Data Preparation**  
Task: Finalize data cleaning and execute feature engineering.  
Yawei XI: Responsible for feature engineering, including applying log transformations to skewed numeric features.  
Yifeng LIN: Responsible for data cleaning.

- **Week 2: Modeling & Optimization**  
Task: Train Random Forest. Conduct hyperparameter tuning and extract feature importance scores.  
Yawei XI: Responsible for conducting hyperparameter tuning and extracting feature importance scores.  
Yifeng LIN: Responsible for building and training the initial Random Forest models.  

- **Week 3: Evaluation & Analysis**  
Task: Evaluate model results using the defined evaluation metrics and generate all necessary performance visualizations.  
Yawei XI: Responsible for computing and analyzing the core evaluation metrics.  
Yifeng LIN: Responsible for generating performance visualizations, specifically the Confusion Matrix for error analysis.  

- **Week 4: Final Deliverables**  
Task: Compile the findings into the final written report and prepare the slide deck for the final presentation.  
Yawei XI: Responsible for drafting, structuring, and formatting the final written report.  
Yifeng LIN: Responsible for designing and preparing the presentation slide deck.  
Both members will conduct a final review.